---
title: "Covid Data Analysis"
author: "Karan Juneja"
date: "19/09/2021"
output: pdf_document
---

We need to analyse the covid data collected by john's hopkins university, and produce some insights on the data.

```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)
library(tidyquant)
library(cowplot)
```

# Loading Files

Source is the https://github.com/CSSEGISandData/COVID-19,
Which is COVID-19 Data Repository by the Center for Systems Science and 
Engineering (CSSE) at Johns Hopkins University.
License : Creative Commons Attribution 4.0 International (CC BY 4.0) by the Johns Hopkins University.

```{r file loading}

url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_"
file_names <- c("confirmed_global.csv",
                    "deaths_global.csv",
                    "confirmed_US.csv",
                    "deaths_US.csv")
urls <- str_c(url_in,file_names)
```

# Reading files

Reading the files using **read_csv**, since files are comma separated. 

```{r reading}
global_cases <- read_csv(urls[1], show_col_types = FALSE)
global_deaths <- read_csv(urls[2], show_col_types = FALSE) 
US_cases <- read_csv(urls[3], show_col_types = FALSE)
US_deaths <- read_csv(urls[4], show_col_types = FALSE)
```

# Preprocessing 

Here we preprocess the data, clean it up converting dates to date formats using lubridate. We also convert the data
to a format that we can use to analyze.

```{r Preprocessing}


global_cases <- global_cases %>%
    pivot_longer(cols = -c(`Province/State`, 
                           `Country/Region`, Lat, Long), 
                 names_to = "date", 
               values_to = "cases") %>%
  select(-c(Lat,Long))

global_deaths <- global_deaths %>%
    pivot_longer(cols = -c(`Province/State`,
                           `Country/Region`, Lat, Long), 
                 names_to = "date", 
               values_to = "deaths") %>%
  select(-c(Lat, Long))


global <- global_cases %>% 
  full_join(global_deaths) %>%
  rename(Country_Region = `Country/Region`, 
         Province_State = `Province/State`) %>%
  mutate(date = mdy(date))

```


Summary of the data

```{r}
summary(global)
```

Removing all the rows on where there was no covid case.

```{r}
global <- global %>% filter(cases > 0)
```


Creating a key to join population data of a country with global dataframe.

```{r}

global <- global %>% 
  unite("Combined_Key", 
                 c(Province_State, Country_Region), 
                                   sep = ", ", 
                 na.rm = TRUE, 
                 remove = FALSE)
```

Loading the population data from the github repository.

```{r}

uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"
uid <- read_csv(uid_lookup_url) %>%
  select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))

```

Joining both the data frames using left join on "Province_State" and "Country_Region".

```{r}

global <- global %>% 
  left_join(uid, by = c("Province_State", "Country_Region")) %>%
  select(-c(UID, FIPS)) %>%
  select(Province_State, Country_Region, date,
         cases, deaths, Population,
         Combined_Key)
global
```

# Preprocessing United States Data

Here we preprocess the data, clean it up converting dates to date formats using lubridate. We also convert the data
to a format that we can use to analyze US data, Also Combining the Deaths data with Cases data and creating a new
dataframe called "US".

```{r}
US_cases <- US_cases %>%
    pivot_longer(cols = -(UID:Combined_Key), 
                 names_to = "date", 
                 values_to = "cases") %>%
    select(Admin2:cases) %>%
    mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))
 
US_deaths <- US_deaths %>%
    pivot_longer(cols = -(UID:Population), 
                 names_to = "date", 
                 values_to ="deaths") %>%
    select(Admin2:deaths) %>%
    mutate(date = mdy(date)) %>%
 select(-c(Lat, Long_))
US <- US_cases %>% 
  full_join(US_deaths)
```


```{r}
US_by_state <- US %>%
  group_by(Province_State, Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths), 
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths *1000000 / Population) %>%
  select(Province_State, Country_Region, date,
         cases, deaths, deaths_per_mill, Population) %>%
  ungroup()

US_totals <- US_by_state %>%
  group_by(Country_Region, date) %>%
   summarize(cases = sum(cases), deaths = sum(deaths), 
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths *1000000 / Population) %>%
  select(Country_Region, date,
         cases, deaths, deaths_per_mill, Population) %>%
  ungroup()

```

## PLOTS

The plot below shows deaths and cases in united states from patient zero to today.

```{r}

US_totals %>% 
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
    geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
    scale_y_log10() +
    theme(legend.position="bottom",
          axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID19 in US", y= NULL)

```

The plot below shows deaths and cases in New york from patient zero to today.

```{r}
state <- "New York"
US_by_state %>% 
  filter(Province_State == state) %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
    geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
    scale_y_log10() +
    theme(legend.position="bottom",
          axis.text.x = element_text(angle = 90)) +
  labs(title = str_c("COVID19 in ", state), y= NULL)
```
if we see the y axis in the graph above the values peak at 1e+05 which is a huge number and doesn't really tell us much,
therefore using the lag() function we will create new columns **new_cases** and **new_deaths** i.e the number of cases
per day and the number of deaths per day.

```{r}
US_by_state <- US_by_state %>%
  mutate(new_cases = cases - lag(cases), 
  new_deaths = deaths - lag(deaths))
US_totals <- US_totals %>%
  mutate(new_cases = cases - lag(cases), 
  new_deaths = deaths - lag(deaths))
```

Now creating a new graph with the columns **new_cases** and **new_deaths**, we can observe the trends in more depth per day.

```{r}
US_totals %>%
  ggplot(aes(x = date, y = new_cases)) +
    geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
    scale_y_log10() +
    theme(legend.position="bottom",
          axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID19 in US", y= NULL)
```
Same graph as above but only for 1 state i.e new york, but still it doesn't show the full story.

```{r}
state <- "New York"
US_by_state %>% 
  filter(Province_State == state) %>%
  ggplot(aes(x = date, y = new_cases)) +
    geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
    scale_y_log10() +
    theme(legend.position="bottom",
          axis.text.x = element_text(angle = 90)) +
  labs(title = str_c("COVID19 in ", state), y= NULL)
```

Below we are finding out the best and worst states in terms of deaths and cases per 1000.
We use the **slice_min** to find the top 10 best states and **slice_max** to find the top 10
worst states in terms of deaths and cases per 1000 people.

```{r}
US_state_totals <- US_by_state %>% 
  group_by(Province_State) %>%
  summarize(deaths = max(deaths), cases = max(cases),
            population = max(Population), 
            cases_per_thou = 1000* cases / population,
            deaths_per_thou = 1000 * deaths / population) %>% 
  filter(cases > 0, population > 0) 

US_state_totals %>%
  slice_min(deaths_per_thou, n = 10)
```


```{r}
US_state_totals %>% 
  slice_max(deaths_per_thou, n = 10)
```
So my interest in the data is how do we find when the covid cases in a country have
peaked and also has the deaths been peaked?

From the total cases and death plots above we really can't see the actual trends such 
as the trends which might tell us when the cases are plateauing, details about the 
what wave are we in?? and which part of the wave are we in etc. So we can use 
**Moving averages** for these analysis.

Below are the plots made using 7 day **simple moving average** and 30 day *SMA* using
the tidyquant package.

We can see from the graph that in the 2nd wave as soon as the 30 day sma plateaued
after that the cases started falling. We can see the same trend happening in the third wave that's 
ongoing right now, but i think we can confidently deduce that the cases are plateauing and we might
see a huge drop in cases soon.

Let's look at the deaths, so in the first wave deaths peaked at 2000 deaths a day, 2nd wave they peaked at 
3000 deaths a day, but right now in the ongoing 3rd wave the deaths still haven't plateaued and that's a
scary concern. But since the cases have started to plateau and also vaccine rollouts, the deaths might not reach
the 2nd wave peak of 3000 deaths a day.

```{r}
cases <- US_totals %>%
  ggplot(aes(x = date, y = new_cases)) +
   geom_ma(ma_fun = SMA, n = 7,alpha = 0.5) +                
  geom_ma(ma_fun = SMA, n = 30, color = "red")+
  labs(title = "SMA Cases per day")
deaths <- US_totals %>% 
  ggplot(aes(x = date, y = new_deaths)) +
  geom_ma(ma_fun = SMA, n = 7,alpha = 0.5) +                 
  geom_ma(ma_fun = SMA, n = 30, color = "red")+
  labs(title = "SMA deaths per day")

plot_grid(cases,deaths,ncol=1,align='v')

```

The state of alabama shows almost the same pattern as the whole of US but, we can see that the 
cases have started to drop and the 7 day sma kind of proves that as well as 30 day sma concludes 
the findings.

But the deaths still are rising in alabama too its a scary thing.

```{r}
state <- "Alabama"
cases <- US_by_state %>% 
  filter(Province_State == state) %>%
  ggplot(aes(x = date, y = new_cases)) +
  geom_ma(ma_fun = EMA, n = 7,alpha=0.5) +                      # Plot 30-day EMA
  geom_ma(ma_fun = EMA, n = 30, color = "red")+
  labs(title = "SMA Cases per day in alabama")

deaths <- US_totals %>% 
  ggplot(aes(x = date, y = new_deaths)) +
  geom_ma(ma_fun = SMA, n = 7,alpha = 0.5) +                 
  geom_ma(ma_fun = SMA, n = 30, color = "red")+
  labs(title = "SMA deaths per day in alabama")

plot_grid(cases,deaths,ncol=1,align='v')

```

## Model

Lets create a linear model with target as *death_per_thou* and predictor as *cases_per_thou*, what
we mean by creating the model is that given the *cases_per_thou* can we predict the  *death_per_thou*,
using a linear regression model.

```{r}
mod <- lm(deaths_per_thou ~ cases_per_thou, data = US_state_totals)
summary(mod)
```

```{r}
x_grid <- seq(1, 151)
new_df <- tibble(cases_per_thou = x_grid)
US_state_totals %>% mutate(pred = predict(mod))
```

```{r}
US_tot_w_pred <- US_state_totals %>% mutate(pred = predict(mod))
US_tot_w_pred %>% ggplot() +
geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "blue") +
geom_point(aes(x = cases_per_thou, y = pred), color = "red")
```

## Bias

+ Dataset Bias

  + How the cases are counted?,if a person takes 3 tests and comes positive 3 times 
  is that also counted as a single case or a multiple cases?
  
+ Personal Bias
  
  + So my bias was that due to vaccinations,there would be less number of cases.But the
  cases in the third wave seem to be peaking at the same level as the 2nd wave when there
  were no vaccinations.



## Conclusion

Covid is not gone, even with the rigorous vaccine rollouts we can see that the number of cases per
day in the 3rd wave seem to have peaked out at the same level as the number of cases per day during the
2nd Wave.

But deaths haven't peaked out, yes at the same time in the 2nd wave at the same number of cases there were
more deaths then now (due to vaccinations?).But the deaths Moving average graph hasn't peaked yet so time will tell
how many lives will be lost due to this pandemic.But looking at the countries such as the UK the third wave even if
it had as many cases per day as the 2nd wave, but it still had way fewer deaths then in the 2nd wave.So lets hope
for the best and Mask Up.